\section*{Editorial and Project Disclaimer}

This volume is an independent scholarly and technical project and is not affiliated
with or endorsed by any church body, academic institution, or organization.

The texts presented herein are historical documents. English translations and
editorial materials are provided for educational and research purposes only.
They do not claim doctrinal, ecclesiastical, or institutional authority.
The original source texts remain definitive.

The English translations in this volume were generated primarily through artificial intelligence–based language tools operating under detailed scholarly instructions. These translations were subsequently reviewed, evaluated, and corrected through an independent secondary AI process for semantic, theological, and stylistic fidelity to the original Norwegian sources. Editorial responsibility in the present edition is limited to methodological design, critical oversight, and presentation; the original source texts remain authoritative.

Please see \texttt{DISCLAIMER.txt} for further details regarding scope,
methodology, and limitations.

Editorial Sigla.—

Notes marked “—Ed.” originate with Andreas Helland, editor of the 1910 volume.

Notes marked “— Present Ed.” indicate editorial responsibility in the present edition.


\subsection*{Translator’s Note on Scriptural Infrastructure and KJV Influence}

A persistent methodological difficulty underlying the present volume must be stated explicitly: the Danish–Lutheran biblical tradition presupposed by the source texts does not possess a full, historically corresponding English counterpart. The Old Testament of the Danish Church (1819) and the Authorized Revision of the New Testament (1871) together form a coherent confessional and rhetorical scriptural register that differs in significant respects from the English King James Version, particularly in matters of verbal aspect, tense orientation, and missionary futurity. No complete English Bible translated directly and consistently from this Danish tradition presently exists.

As a result, the present translations necessarily operate within a structural constraint. While the project intentionally adopts a policy of formal equivalence oriented toward Danish–Lutheran semantics and theology, extended biblical quotations and densely scriptural passages in English inevitably risk absorbing syntactic, idiomatic, or phrasal elements from the King James Bible. This phenomenon should be understood not as a conscious theological harmonization toward the KJV tradition, but as a form of linguistic “bleeding through” arising from the absence of an established English sacred register corresponding to the Danish Bible used by the original authors.

In other words, the AI is not merely translating Danish theological prose into English; they are translating prose that already presupposes a biblical language which English does not, in a strict sense, possess. Where English biblical diction is required at scale, the King James Version remains the dominant inherited idiom available to English readers, and its influence cannot be entirely neutralized without the creation of an entirely new English translation of the Danish Bible itself.

Accordingly, instances of residual KJV phrasing should be read as provisional accommodations to this infrastructural gap rather than as corrections, normalizations, or theological realignments. The Danish–Lutheran Scriptures remain normative for interpretation and intent; the English renderings remain approximations constrained by the limits of the target language. This tension is openly acknowledged and left visible, not concealed, in order to preserve historical honesty and scholarly transparency.

The present volume therefore does not claim to resolve this difficulty, but to document it faithfully. Any future project seeking to eliminate such interference would necessarily require the prior production of a comprehensive English translation of the Danish Bible (1819/1871), capable of supplying a stable, non-KJV biblical register for use in historical and theological translation work.